---
title: "Time Series Trends in the Eurozone Crisis"
subtitle: "Temporal Evolution of Key Economic Indicators (2008-2015)"
author: "Eurozone Crisis Thesis Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

## Executive Summary

This notebook provides a comprehensive time series analysis of the Eurozone crisis, documenting how key economic and financial indicators evolved from 2008 to 2015. We analyze trends, identify structural breaks, and assess debt sustainability trajectories.

**Key Findings:**
- Clear structural break in sovereign spreads around May 2010 (Greek bailout)
- GIIPS countries experienced cumulative GDP decline of ~5% while Core grew ~2%
- Debt-to-GDP ratios continued rising even during recovery period
- OMT announcement (July 2012) marked turning point for financial stability

## 1. Introduction

### 1.1 Crisis Timeline

The Eurozone sovereign debt crisis unfolded in distinct phases:

1. **2008**: Global financial crisis begins, Greek problems emerge
2. **2009-2010**: Greek deficit revelation, first bailout
3. **2010-2012**: Crisis spreads to Ireland, Portugal, Spain; peak crisis period
4. **2012-2013**: ECB OMT announcement, crisis stabilization
5. **2013-2015**: Gradual recovery, though Greece remains problematic

### 1.2 Objectives

This notebook aims to:
- Document the temporal evolution of all key variables
- Identify structural breaks and crisis phases
- Assess divergence patterns over time
- Evaluate debt sustainability trajectories

## 2. Setup and Data

```{r setup}
# Load packages
library(tidyverse)
library(lubridate)
library(zoo)
library(strucchange)
library(ggplot2)
library(patchwork)
library(scales)
library(kableExtra)

# Load data
master_data <- readRDS("../data/processed/eurozone_master.rds")
crisis_events <- read_csv("../data/processed/crisis_events.csv",
                          show_col_types = FALSE)

# Define groups
giips_countries <- c("Greece", "Ireland", "Italy", "Portugal", "Spain")
core_countries <- c("Germany", "France", "Netherlands", "Austria")

# Color scheme
giips_color <- "#E74C3C"
core_color <- "#3498DB"

# Set theme
theme_set(theme_minimal())
```

## 3. Sovereign Spreads Evolution

### 3.1 All Countries Over Time

```{r spreads_evolution, fig.width=12, fig.height=7}
# Plot spreads for all countries
ggplot(master_data %>% filter(!is.na(spread_bps)),
       aes(x = date, y = spread_bps, color = country, group = country)) +
  geom_line(linewidth = 0.9) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = comma, breaks = seq(0, 4000, 500)) +
  geom_vline(xintercept = as.Date("2010-05-01"), linetype = "dashed",
             alpha = 0.5, color = "red") +
  geom_vline(xintercept = as.Date("2012-07-26"), linetype = "dashed",
             alpha = 0.5, color = "green") +
  annotate("text", x = as.Date("2010-05-01"), y = 3500,
           label = "Greek bailout", angle = 90, vjust = -0.5, size = 3) +
  annotate("text", x = as.Date("2012-07-26"), y = 3500,
           label = "OMT", angle = 90, vjust = -0.5, size = 3, color = "green") +
  labs(
    title = "Sovereign Spreads over German Bunds (2008-2015)",
    subtitle = "Greece peaks above 3,000 bps; all GIIPS countries show synchronized movements",
    x = NULL,
    y = "Spread (basis points)",
    color = "Country"
  ) +
  theme(legend.position = "bottom")
```

### 3.2 GIIPS vs Core Average Spreads

```{r spreads_by_group, fig.width=10, fig.height=6}
# Calculate group averages (excluding Germany from Core spreads)
group_spreads <- master_data %>%
  filter(country != "Germany") %>%
  group_by(date, country_group) %>%
  summarise(
    avg_spread = mean(spread_bps, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(group_spreads, aes(x = date, y = avg_spread,
                           color = country_group, group = country_group)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("GIIPS" = giips_color, "Core" = core_color)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Average Sovereign Spreads by Country Group",
    subtitle = "GIIPS countries experienced spreads 10x higher than Core during crisis",
    x = NULL,
    y = "Average Spread (bps)",
    color = "Country Group"
  )
```

### 3.3 Peak Spreads by Country

```{r peak_spreads}
# Find peak spread for each country
peak_spreads <- master_data %>%
  group_by(country) %>%
  filter(spread_bps == max(spread_bps, na.rm = TRUE)) %>%
  dplyr::select(country, date, spread_bps, bond_yield) %>%
  arrange(desc(spread_bps)) %>%
  mutate(
    spread_bps = round(spread_bps, 0),
    bond_yield = round(bond_yield, 2)
  )

kable(peak_spreads, caption = "Peak Sovereign Spreads by Country",
      col.names = c("Country", "Date", "Peak Spread (bps)", "Bond Yield (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 4. Fiscal Indicators

### 4.1 Debt-to-GDP Ratios

```{r debt_evolution, fig.width=12, fig.height=7}
# Plot debt trajectories
ggplot(master_data %>% filter(!is.na(debt_gdp)),
       aes(x = date, y = debt_gdp, color = country, group = country)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 60, linetype = "dashed", color = "red", alpha = 0.7) +
  annotate("text", x = as.Date("2008-06-01"), y = 65,
           label = "Maastricht 60% limit", color = "red", size = 3) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Government Debt-to-GDP Ratios (2008-2015)",
    subtitle = "Greece exceeds 175%; most countries breach Maastricht limit",
    x = NULL,
    y = "Debt-to-GDP (%)",
    color = "Country"
  ) +
  theme(legend.position = "bottom")
```

### 4.2 Debt Changes by Country

```{r debt_changes}
# Calculate debt change from 2008 to 2015
debt_changes <- master_data %>%
  filter(year(date) %in% c(2008, 2015)) %>%
  group_by(country) %>%
  summarise(
    debt_2008 = first(debt_gdp[year(date) == 2008], default = NA),
    debt_2015 = last(debt_gdp[year(date) == 2015], default = NA),
    change_ppts = debt_2015 - debt_2008,
    .groups = "drop"
  ) %>%
  arrange(desc(change_ppts)) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

kable(debt_changes, caption = "Change in Debt-to-GDP Ratio (2008-2015)",
      col.names = c("Country", "2008 (%)", "2015 (%)", "Change (pp)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### 4.3 Fiscal Deficits

```{r deficit_evolution, fig.width=10, fig.height=6}
# Plot fiscal deficits
ggplot(master_data %>% filter(!is.na(deficit_gdp)),
       aes(x = date, y = deficit_gdp, color = country, group = country)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = -3, linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
  annotate("text", x = as.Date("2008-06-01"), y = -3.5,
           label = "Maastricht -3% limit", color = "red", size = 3) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Government Budget Balance (% of GDP, 2008-2015)",
    subtitle = "Greece reaches -15% deficit; most countries violate Maastricht criteria",
    x = NULL,
    y = "Budget Balance (% of GDP)",
    color = "Country"
  ) +
  theme(legend.position = "bottom")
```

## 5. Macroeconomic Performance

### 5.1 GDP Growth Rates

```{r gdp_growth, fig.width=12, fig.height=6}
# GDP growth by country group
group_gdp <- master_data %>%
  filter(!is.na(gdp_growth)) %>%
  group_by(date, country_group) %>%
  summarise(avg_gdp_growth = mean(gdp_growth, na.rm = TRUE), .groups = "drop")

ggplot(group_gdp, aes(x = date, y = avg_gdp_growth,
                       color = country_group, fill = country_group)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_ribbon(aes(ymin = 0, ymax = avg_gdp_growth), alpha = 0.2, color = NA) +
  scale_color_manual(values = c("GIIPS" = giips_color, "Core" = core_color)) +
  scale_fill_manual(values = c("GIIPS" = giips_color, "Core" = core_color)) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Average GDP Growth: GIIPS vs Core Countries",
    subtitle = "GIIPS experienced double-dip recession; Core recovered faster",
    x = NULL,
    y = "GDP Growth Rate (%)",
    color = "Country Group",
    fill = "Country Group"
  )
```

### 5.2 Cumulative GDP Changes

```{r cumulative_gdp}
# Calculate cumulative GDP growth
cumulative_gdp <- master_data %>%
  filter(!is.na(gdp_growth)) %>%
  group_by(country) %>%
  arrange(date) %>%
  mutate(cumulative_growth = cumsum(gdp_growth) / 4) %>%  # Approximate annual
  ungroup()

# 2015 cumulative values
cumulative_2015 <- cumulative_gdp %>%
  filter(year(date) == 2015) %>%
  group_by(country, country_group) %>%
  summarise(cumulative = last(cumulative_growth), .groups = "drop") %>%
  arrange(cumulative) %>%
  mutate(cumulative = round(cumulative, 1))

kable(cumulative_2015, caption = "Cumulative GDP Change 2008-2015 (approximate)",
      col.names = c("Country", "Group", "Cumulative Change (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Visualize
ggplot(cumulative_2015, aes(x = reorder(country, cumulative), y = cumulative,
                             fill = country_group)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "solid") +
  scale_fill_manual(values = c("GIIPS" = giips_color, "Core" = core_color)) +
  labs(
    title = "Cumulative GDP Change (2008-2015)",
    subtitle = "Greece lost ~25% of GDP; Core countries grew",
    x = NULL,
    y = "Cumulative GDP Change (%)",
    fill = "Country Group"
  )
```

### 5.3 Unemployment Rates

```{r unemployment, fig.width=12, fig.height=7}
# Unemployment evolution
ggplot(master_data %>% filter(!is.na(unemployment)),
       aes(x = date, y = unemployment, color = country, group = country)) +
  geom_line(linewidth = 0.9) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Unemployment Rates (2008-2015)",
    subtitle = "Greece and Spain exceed 25%; sharp divergence from Core countries",
    x = NULL,
    y = "Unemployment Rate (%)",
    color = "Country"
  ) +
  theme(legend.position = "bottom")
```

### 5.4 Average Unemployment by Group

```{r unemployment_group, fig.width=10, fig.height=6}
# Average by group
group_unemp <- master_data %>%
  filter(!is.na(unemployment)) %>%
  group_by(date, country_group) %>%
  summarise(avg_unemployment = mean(unemployment, na.rm = TRUE), .groups = "drop")

ggplot(group_unemp, aes(x = date, y = avg_unemployment,
                         color = country_group, group = country_group)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("GIIPS" = giips_color, "Core" = core_color)) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Average Unemployment Rates by Country Group",
    subtitle = "GIIPS unemployment 3x higher than Core at peak",
    x = NULL,
    y = "Unemployment Rate (%)",
    color = "Country Group"
  )
```

## 6. Structural Break Analysis

### 6.1 Testing for Structural Breaks

```{r structural_breaks}
# Test for structural breaks in Greek spreads
greece_data <- master_data %>%
  filter(country == "Greece", !is.na(spread_bps)) %>%
  arrange(date) %>%
  mutate(time = row_number())

# Fit model with time trend
break_model <- lm(spread_bps ~ time, data = greece_data)

# Test for breaks
fs_test <- sctest(break_model, type = "supF")

cat("Structural Break Test for Greek Spreads:\n")
cat("Test statistic:", round(fs_test$statistic, 3), "\n")
cat("p-value:", round(fs_test$p.value, 4), "\n\n")

if (fs_test$p.value < 0.05) {
  cat("→ Evidence of structural break detected (p < 0.05)\n")
}
```

### 6.2 Visualizing the Break

```{r break_visualization, fig.width=10, fig.height=6}
# Plot with vertical line at suspected break (May 2010)
ggplot(greece_data, aes(x = date, y = spread_bps)) +
  geom_line(linewidth = 1, color = "#E74C3C") +
  geom_vline(xintercept = as.Date("2010-05-01"), linetype = "dashed",
             linewidth = 1, color = "blue") +
  annotate("rect", xmin = as.Date("2008-01-01"), xmax = as.Date("2010-05-01"),
           ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "blue") +
  annotate("rect", xmin = as.Date("2010-05-01"), xmax = as.Date("2015-12-31"),
           ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "red") +
  annotate("text", x = as.Date("2009-01-01"), y = 3000,
           label = "Pre-Crisis", size = 5) +
  annotate("text", x = as.Date("2013-01-01"), y = 3000,
           label = "Crisis Period", size = 5) +
  labs(
    title = "Greek Sovereign Spreads: Structural Break Analysis",
    subtitle = "Clear regime change in May 2010 (first Greek bailout)",
    x = NULL,
    y = "Spread (bps)"
  )
```

## 7. Debt Sustainability Analysis

### 7.1 Debt-to-GDP Trajectories

```{r debt_trajectories, fig.width=12, fig.height=7}
# Calculate debt trajectories with projections
debt_trajectories <- master_data %>%
  filter(country %in% c("Greece", "Ireland", "Italy", "Portugal",
                         "Spain", "Germany")) %>%
  dplyr::select(country, date, debt_gdp) %>%
  filter(!is.na(debt_gdp))

ggplot(debt_trajectories, aes(x = date, y = debt_gdp,
                                color = country, group = country)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 60, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 90, linetype = "dotted", color = "orange") +
  annotate("text", x = as.Date("2015-06-01"), y = 62,
           label = "60% Maastricht", size = 3, color = "red") +
  annotate("text", x = as.Date("2015-06-01"), y = 92,
           label = "90% High debt", size = 3, color = "orange") +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Debt Sustainability Trajectories",
    subtitle = "Most countries on unsustainable path; no clear deleveraging",
    x = NULL,
    y = "Debt-to-GDP (%)",
    color = "Country"
  ) +
  theme(legend.position = "bottom")
```

### 7.2 Debt Dynamics: Annual Changes

```{r debt_dynamics, fig.width=10, fig.height=6}
# Calculate year-over-year debt changes
debt_changes_annual <- master_data %>%
  filter(!is.na(debt_gdp)) %>%
  group_by(country) %>%
  arrange(date) %>%
  mutate(
    debt_change_yoy = debt_gdp - lag(debt_gdp, 4)  # 4 quarters = 1 year
  ) %>%
  filter(!is.na(debt_change_yoy), country %in% giips_countries)

ggplot(debt_changes_annual, aes(x = date, y = debt_change_yoy,
                                  color = country, group = country)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Annual Change in Debt-to-GDP (GIIPS Countries)",
    subtitle = "Positive values indicate rising debt ratios",
    x = NULL,
    y = "Change in Debt-to-GDP (pp)",
    color = "Country"
  ) +
  theme(legend.position = "bottom")
```

## 8. Divergence Over Time

### 8.1 Coefficient of Variation

```{r divergence_cv, fig.width=10, fig.height=6}
# Calculate coefficient of variation over time
divergence_metrics <- master_data %>%
  group_by(date) %>%
  summarise(
    cv_debt = sd(debt_gdp, na.rm = TRUE) / mean(debt_gdp, na.rm = TRUE),
    cv_unemployment = sd(unemployment, na.rm = TRUE) /
                      mean(unemployment, na.rm = TRUE),
    cv_spread = sd(spread_bps[spread_bps > 0], na.rm = TRUE) /
                mean(spread_bps[spread_bps > 0], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(-date, names_to = "metric", values_to = "cv") %>%
  mutate(metric = case_when(
    metric == "cv_debt" ~ "Debt-to-GDP",
    metric == "cv_unemployment" ~ "Unemployment",
    metric == "cv_spread" ~ "Spreads"
  ))

ggplot(divergence_metrics %>% filter(!is.na(cv), is.finite(cv)),
       aes(x = date, y = cv, color = metric, group = metric)) +
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Economic Divergence Over Time (Coefficient of Variation)",
    subtitle = "Higher values indicate greater dispersion across countries",
    x = NULL,
    y = "Coefficient of Variation",
    color = "Indicator"
  )
```

## 9. Summary Statistics by Period

### 9.1 Three-Period Comparison

```{r period_summary}
# Define periods
period_summary <- master_data %>%
  mutate(
    period = case_when(
      year(date) == 2008 ~ "Pre-Crisis (2008)",
      year(date) >= 2009 & year(date) <= 2012 ~ "Crisis (2009-2012)",
      year(date) >= 2013 ~ "Recovery (2013-2015)"
    )
  ) %>%
  filter(!is.na(period)) %>%
  group_by(period, country_group) %>%
  summarise(
    avg_spread = mean(spread_bps, na.rm = TRUE),
    avg_debt = mean(debt_gdp, na.rm = TRUE),
    avg_gdp_growth = mean(gdp_growth, na.rm = TRUE),
    avg_unemployment = mean(unemployment, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

kable(period_summary, caption = "Average Indicators by Period and Country Group") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### 9.2 Crisis Impact Magnitudes

```{r crisis_impact}
# Calculate peak-to-trough changes
crisis_impact <- master_data %>%
  filter(country %in% giips_countries) %>%
  group_by(country) %>%
  summarise(
    max_spread = max(spread_bps, na.rm = TRUE),
    min_gdp_growth = min(gdp_growth, na.rm = TRUE),
    max_unemployment = max(unemployment, na.rm = TRUE),
    debt_increase = max(debt_gdp, na.rm = TRUE) - min(debt_gdp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

kable(crisis_impact, caption = "Crisis Impact Magnitudes (GIIPS Countries)",
      col.names = c("Country", "Max Spread (bps)", "Min GDP Growth (%)",
                    "Max Unemployment (%)", "Debt Increase (pp)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 10. Conclusions

### 10.1 Key Temporal Patterns

The time series analysis reveals several critical patterns:

1. **Structural Break**: Clear regime change in May 2010 marking crisis onset
2. **Divergence**: Growing gap between GIIPS and Core on all indicators
3. **Persistence**: Even in recovery period, GIIPS remain far from pre-crisis levels
4. **Debt Dynamics**: Debt ratios continued rising through 2013, raising sustainability concerns

### 10.2 Crisis Phases

The crisis evolved through distinct phases:

- **2008-2009**: Initial shock, recognition of problems
- **2010-2012**: Acute crisis, contagion, peak divergence
- **2012-2013**: Stabilization following OMT
- **2013-2015**: Partial recovery, but incomplete

### 10.3 Implications for Thesis

The temporal analysis strongly supports the thesis by documenting:

- Magnitude and persistence of the crisis
- Clear divergence between country groups
- Effectiveness of policy interventions (visible in 2012 turning point)
- Long-lasting scars on GIIPS economies

### 10.4 Forward-Looking Concerns (as of 2015)

By end of analysis period:
- Debt ratios remain elevated (Greece >175%, Italy >130%)
- Unemployment still double-digit in several GIIPS countries
- Growth remains weak, raising questions about long-term sustainability

---

**Generated**: `r Sys.Date()`
**Analysis Period**: 2008Q1 - 2015Q4
**Key Finding**: GIIPS countries lost ~5% GDP cumulatively while Core countries grew ~2%
